Highly Available Transactions: Virtues and Limitations

### 2013-03-31 writeup/motivation.tex
PDB pass plus Alan comments

\section{Partitions and Latency}
\label{sec:motivation}

Distributed databases face a host of challenges that their
single-server counterparts are exempted from, largely stemming from
network asynchrony and partial failure. These challenges mean that
maintaining semantic guarantees has a cost, the exact price of which
is a subject of considerable speculation in the database
community~\cite{stonebraker2010errors}. In this section, we attempt to
quantify the prevalence and cost of network partitions and latency in
several real-world deployments. This data and a recent survey of
existing database semantics motivates the study of highly available
transaction systems.

(TODO-PDB: keep this?) We will begin by studying the behavior of
network partitions and latency. Peter Deutsch begins his classic list
of ``Fallacies of Distributed Computing'' with two concerns
fundamental to distributed database systems: ``\textit{1.)}  The
network is reliable. \textit{2.)} Latency is
zero''~\cite{fallacies-deutsch}. On a single server, communication
channels are robust and relatively fast, and few single-server systems
need to account for a possibility of partial system failure. However,
in a distributed setting, both of these assumptions are no longer
valid. To extend Jim Gray's storage latency analogy~\cite{gray-rules},
if memory is 1.5 hours from the CPU and disk is 2 years away, then,
especially across wide-area networks (WANs), remote server accesses
can exceed 30 years, with the added concern that messages can be lost
in transit or delayed indefinitely by network partitions. Accordingly,
ignoring network failures and latency can ``cause big trouble and
painful learning experiences''~\cite{fallacies-deutsch}.

\subsection{Partitions in the Real World}

Network partitions---informally, communication failures between
participants in a distributed system---are challenging. A system
facing network partitions faces the impossibility of simultaneously
maintaining a correct ``global'' view of system state while
simultaneously providing online operation to all of its
clients~\cite{davidson-survey}. For networks that do not partition,
maintaining semantic guarantees and providing ``always on'' operation
is straightforward~\cite{stonebraker2010errors}. However, given the
\textit{possibility} of partitions, a system must be prepared to
either lose availability of operations or its ability to reason about
global state. In this section, we present evidence for the frequency
and non-triviality of WAN and LAN partitions.

@@

Several recent studies quantify partition behavior more rigorously. A
2011 study of several Microsoft datacenters found a mean of 40.8
network link failures per day (95th percentile: 136), with a median
time to repair of around five minutes (and up to one week). Perhaps
surprisingly, provisioning redundant networks only reduces impact of
failures by up to 40\%, meaning network providers cannot easily
curtail partition behavior~\cite{sigcomm-dc}. A 2010 study of over 200
wide-area routers found an average of 16.2--302.0 failures per link
per year with an average annual downtime of 24--497 minutes per link
per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
HP's managed enterprise networks, WAN, LAN, and connectivity problems
account for 28.1\% of all customer support tickets while 39\% of
tickets relate to network hardware.  The median incident duration for
highest priority tickets ranges from 114--188 minutes and up to a full
day for all tickets~\cite{turner2012failure}. Other studies confirm
these results, showing median time between failures over a WAN network
of approximately 3000 seconds with a median time to repair between 2
and 1000 seconds~\cite{ip-backbone-failures} as well as frequent path
routing failures on the Internet~\cite{labovitz-failures}. Isolating,
quantifying, and designing for these network failures is an area of
active research in networking
community~\cite{surviving-failures-bodik, uw-failure-networks}.

@@

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' In this
section, we quantify expected latencies, which are often
huge---hundreds of milliseconds---in a geo-replicated context.

Fundamentally, the speed at which two servers can communicate is
(according to modern physics) bounded by the speed of light. Two
servers located on opposite sides of the Earth face, in the best
case---communicating via a hypothetical link crossing through the
center of the planet---require a minimum 85.1 ms RTT (133.7 ms if sent
at ground level). As services are replicated to multiple,
geographically distinct sites, the cost of communication increases to
this lower bound. In real deployments, transit times are actually
slower than the speed of light due to routing, congestion, and
processing times. To illustrate the difference between
intra-datacenter, inter-datacenter, and inter-planetary networks, we
performed a measurement study of network behavior on Amazon's EC2, a
widely-used public compute cloud. We measured one week of ping times
between all seven EC2 geographic ``regions'', across three
``availability zones'' (closely co-located datacenters), and within a
single ``availabilty zone'' (datacenter), at a granularity of 1s
(dataset to be released).

We summarize the results of our network measurement study in
Table~\ref{table:rtt}. On average, intra-datacenter communication (1c)
is between 1.82 and 6.38 times faster than across geographically
co-located datacenters (1b) and between 40 and 647 times faster than
across geographically distributed datacenters (1a). The cost of
wide-area communication exceeds the speed of light: for example,
communicating from S\~{a}o Paulo to Singapore is lower-bounded by
speed-of-light transit RTT of 106.7ms but, on average, incurs a
362.8ms RTT (95th percentile: 649ms). As shown in
Figure~\ref{fig:rtt}, the distribution of latencies varies between
links, but the trend---in line with other recent studies~\cite{mdcc,
  redblue}---is clear: remote communication has a substantial cost.

@@

\begin{table}
\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
IR & & & &  & 346.2   & 239.8   & 234.1  \\
SY & & & & &  & 333.6   & 243.1  \\
SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
\hline
  \end{tabular}
}\vspace{.5em}

\subfloat[Across \texttt{us-east} AZs]{
  \makebox[.2\textwidth]{
    \begin{tabular}{|c|c|c|}\hline
 & \multicolumn{1}{c}{C} & \multicolumn{1}{c|}{D}\\\hline
B & \colorbox{min-lat-color}{1.08} & 3.12 \\
C & & \colorbox{max-lat-color}{3.57}  \\
\hline
  \end{tabular}}
 }
\subfloat[Within \texttt{us-east-a} AZ] {
  \makebox[.25\textwidth]{
  \begin{tabular}{|c|c|c|}\hline
 & \multicolumn{1}{c}{H2} & \multicolumn{1}{c|}{H3}\\\hline
H1  & 0.55   & \colorbox{max-lat-color}{0.56} \\
H2 &  & \colorbox{min-lat-color}{0.50}  \\
\hline
  \end{tabular}
 }}
\caption{Mean RTT times on EC2 (min and max highlighted)}
\label{table:rtt}
\end{table}

@@

\section{CAP and Modern Systems}
\label{sec:modernacid}

Systems designers must make fundamental trade-offs between latency,
system behavior during partitions, and achievable semantics. In this
section, we briefly discuss known semantic limits for systems wishing
to maintain high availability and their impact on modern database
systems. We show that many database systems operate under weak model
whose availability characteristics are unknown, motivating further
study of these models in a highly available context in the remainder
of this paper.

One of the most popularly celebrated impossibility results in
distributed systems is Brewer's CAP Theorem~\cite{brewer-slides},
which, as formally proven by Gilbert and Lynch~\cite{gilbert-cap},
states that it is impossible to maintain linearizability , or the
ability to read the last completed write, (``C,'' or
``consistency''---albeit poorly named) and replica availability, or a
guaranteed response, (``A'') in the presence of network partitions
(``P''). Brewer's Theorem has roots in decades of distributed database
research~\cite{davidson-survey} and has had a substantial impact on
recent large-scale distributed systems design~\cite{bigtable, pnuts,
  dynamo}: while CAP is narrowly scoped to the impossibility of
building highly available linearizable systems, it is often
interpreted to preclude a larger set of guarantees, including ACID
(transactional atomicity, consistency, isolation and durability
guarantees). We are unlikely to determine the exact cause of this
confusion, but it appears as early as the first public announcement of
Brewer's result, in which ''distributed databases'' appear as an
example of unavailable systems~\cite{brewer-slides}. While, in the
same document, Brewer is careful to note that consistency and
availability are a ``spectrum,'' all of ACID is still often labeled
with a binary categorization of ``unavailable''~\cite{hn,
  foundation-article}.

Database researchers and designers have long realized that
serializability---the gold standard of ACID isolation, which
guarantees application-level consistency---is not achievable in a
highly available system~\cite{davidson-survey}.  However, to increase
concurrency, database systems offer a range of ACID properties weaker
than serializability. Within a single-node database, the coordination
penalties associated with serializability can be severe and manifest
themselves in the form of decreased concurrency (and, subsequently,
performance degradation, scalability limitations, and, often, external
aborts like deadlocks)~\cite{gray-isolation}. Instead, databases offer
a host of so-called \textit{weak isolation} models that allow varying
restrictions on the space of schedules that are allowable by the
system~\cite{adya}. None of these weak isolation models guarantees
serializability, but, as we see below, the benefits of these models
are often considered to outweigh costs of possible consistency
anomalies that might arise from their use.

To understand the prevalence of these weak isolation models, we
recently surveyed the default and maximum isolation models provided by
18 databases, often claiming to provide ``ACID'' or
``NewSQL'' functionality~\cite{hat-hotos}. As shown in
Table~\ref{table:existing}, only three out of 18 databases provided
serializability by default, and at least eight did not provide
serializability as an option at all. This is particularly surprising
when we consider the widespread deployment of many of these
non-serializable databases, like Oracle 11g, which are known to power
major businesses and product functionality. Given that these
transactional models are frequently used, our inability to provide
serializability in arbitrary HATs appears non-fatal for practical
applications. If application writers and database vendors have already
decided that the benefits of weak isolation outweigh potential
application inconsistencies, then, in a highly available environment
that prohibits serializability, similar decisions may be tenable.

\begin{table}
\begin{center}
\begin{small}
\begin{tabular}{|l|c|c|}
\hline
Database & Default & Maximum\\\hline
Actian Ingres 10.0/10S & S & S\\
Aerospike & RC & RC\\
Akiban Persistit & SI & SI\\
Clustrix CLX 4100 & RR & ?\\
Greenplum 4.1 & RC & S \\
IBM DB2 10 for z/OS & CS & S\\
IBM Informix 11.50 & Depends & RR\\
MySQL 5.6 & RR & S \\
MemSQL 1b & RC & RC\\
MS SQL Server 2012 & RC & S \\
NuoDB & CR & CR\\
Oracle 11g & RC & SI\\
Oracle Berkeley DB & S & S\\
Oracle Berkeley DB JE & RR & S\\
Postgres 9.2.2 & RC & S\\
SAP HANA & RC & SI\\
ScaleDB 1.02 & RC & RC\\
VoltDB & S & S\\
\hline
\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline

@@

It is unknown \textit{which} of these ACID (particularly isolation)
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically designed for a
single-node context and are, to the best of our knowledge, unavailable
due to a reliance on concurrency control mechanisms like locking that
are not resilient to partial failure
(Section~\ref{sec:evaluation}). Morever, we are not aware of any prior
literature that provides guidance as to the relationship between weak
isolation and high availability: prior work has examined the
relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a
single server~\cite{adya, ansicritique, gray-isolation} but never weak isolation and
high availability taken together.

>>>>

\section{Why Availability?}
\label{sec:motivation}

Why does availability matter? Peter Deutsch starts his classic list of
``Fallacies of Distributed Computing'' with two concerns fundamental
to distributed database systems: ``\textit{1.)}  The network is
reliable. \textit{2.)} Latency is zero''~\cite{fallacies-deutsch}; in
a distributed setting, network failures may prevent database servers
from communicating, and, in the absence of failures, communication is
slowed by factors like physical distance and, more often, congestion
and routing. As we will see (Section~\ref{sec:availability}), highly
available systems minimize the effect of these network
\textit{partitions} and \textit{latency}. In this Section, we draw on
a range of evidence that indicates that partitions occur relatively
frequently in real-world deployments and best-case latencies between
datacenters are significant.

\subsection{Network Partitions at Scale}

@@

Several recent studies rigorously quantify partition behavior. A 2011
study of several Microsoft datacenters found a mean of 40.8 network
link failures per day (95th percentile: 136), with a median time to
repair of around five minutes (and up to one week). Perhaps
surprisingly, provisioning redundant networks only reduces impact of
failures by up to 40\%, meaning network providers cannot easily
curtail partition behavior~\cite{sigcomm-dc}. A 2010 study of over 200
wide-area routers found an average of 16.2--302.0 failures per link
per year with an average annual downtime of 24--497 minutes per link
per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
HP's managed enterprise networks, WAN, LAN, and connectivity problems
account for 28.1\% of all customer support tickets while 39\% of
tickets relate to network hardware.  The median incident duration for
highest priority tickets ranges from 114--188 minutes and up to a full
day for all tickets~\cite{turner2012failure}. Other studies confirm
these results, showing median time between failures over a WAN network
of approximately 3000 seconds with a median time to repair between 2
and 1000 seconds~\cite{ip-backbone-failures} as well as frequent path
routing failures on the Internet~\cite{labovitz-failures}. Isolating,
quantifying, and designing for these network failures is an area of
active research in networking
community~\cite{surviving-failures-bodik, uw-failure-networks}.

@@

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' In this
section, we quantify round-trip latencies, which are often
large---hundreds of milliseconds---in a geo-replicated context.
Fundamentally, the speed at which two servers can communicate is
(according to modern physics) bounded by the speed of light. Two
servers located on opposite sides of the Earth face, in the best
case---communicating via a hypothetical link crossing through the
center of the planet---require a minimum 85.1 ms RTT (133.7 ms if sent
at ground level). As services are replicated to multiple,
geographically distinct sites, the cost of communication increases to
this lower bound.

@@

\begin{table}
\vspace{.5em}

\subfloat[Across \texttt{us-east} AZs]{
  \makebox[.2\textwidth]{
    \begin{tabular}{|c|c|c|}\hline
 & \multicolumn{1}{c}{C} & \multicolumn{1}{c|}{D}\\\hline
B & \colorbox{min-lat-color}{1.08} & 3.12 \\
C & & \colorbox{max-lat-color}{3.57}  \\
\hline
  \end{tabular}}
 }
\subfloat[Within \texttt{us-east-a} AZ] {
  \makebox[.25\textwidth]{
  \begin{tabular}{|c|c|c|}\hline
 & \multicolumn{1}{c}{H2} & \multicolumn{1}{c|}{H3}\\\hline
H1  & 0.55   & \colorbox{max-lat-color}{0.56} \\
H2 &  & \colorbox{min-lat-color}{0.50}  \\
\hline
  \end{tabular}
 }}

\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
IR & & & &  & 346.2   & 239.8   & 234.1  \\
SY & & & & &  & 333.6   & 243.1  \\
SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
\hline
  \end{tabular}
}

\caption{Mean RTT times on EC2 (min and max highlighted)}
\label{table:rtt}
\end{table}

@@

In real deployments, transit times are actually slower than the speed
of light due to routing, congestion, and processing times. To
illustrate the difference between intra-datacenter, inter-datacenter,
and inter-planetary networks, we performed a measurement study of
network behavior on Amazon's EC2, a widely-used public compute
cloud. We measured one week of ping times between all seven EC2
geographic ``regions'', across three ``availability zones'' (closely
co-located datacenters), and within a single ``availabilty zone''
(datacenter), at a granularity of 1s (dataset to be released). We
summarize the results of our network measurement study in
Table~\ref{table:rtt}. On average, intra-datacenter communication (1a)
is between 1.82 and 6.38 times faster than across geographically
co-located datacenters (1b) and between 40 and 647 times faster than
across geographically distributed datacenters (1c). The cost of
wide-area communication exceeds the speed of light: for example,
communicating from S\~{a}o Paulo to Singapore is lower-bounded by
speed-of-light transit RTT of 106.7ms but, on average, incurs a
362.8ms RTT (95th percentile: 649ms). As shown in
Figure~\ref{fig:rtt}, the distribution of latencies varies between
links, but the trend---in line with other recent studies~\cite{mdcc,
  redblue}---is clear: remote communication has a substantial cost.

\section{When is ``ACID'' ACID?}
\label{sec:modernacid}

The previous section demonstrated that distributed systems must
address partitions and latency: what does this mean for distributed
databases? Database researchers and designers have long realized that
serializability is not achievable in a highly available
system~\cite{davidson-survey}, meaning it will be expensive in the
environments from Section~\ref{sec:motivation}. However, even a
single-node database, the coordination penalties associated with
serializability can be severe, manifesting themselves in the form of
decreased concurrency (and, subsequently, performance degradation,
scalability limitations, and, often, external aborts like
deadlocks)~\cite{gray-isolation}. Accordingly, o increase concurrency,
database systems offer a range of ACID properties weaker than
serializability.  Instead, databases offer a host of so-called
\textit{weak isolation} models that allow varying restrictions on the
space of schedules that are allowable by the system~\cite{adya,
  ansicritique, ansi-sql}. None of these weak isolation models
guarantees serializability, but, as we see below, the benefits of
these models are often considered to outweigh costs of possible
consistency anomalies that might arise from their use.

To understand the prevalence of these weak isolation models, we
recently surveyed the default and maximum isolation models provided by
18 databases, often claiming to provide ``ACID'' or ``NewSQL''
functionality~\cite{hat-hotos}. As shown in
Table~\ref{table:existing}, only three out of 18 databases provided
serializability by default, and eight did not provide serializability
as an option at all. This is particularly surprising when we consider
the widespread deployment of many of these non-serializable databases,
like Oracle 11g, which are known to power major businesses and product
functionality. Given that these transactional models are frequently
used, our inability to provide serializability in arbitrary HATs
appears non-fatal for practical applications. If application writers
and database vendors have already decided that the benefits of weak
isolation outweigh potential application inconsistencies, then, in a
highly available environment that prohibits serializability, similar
decisions may be tenable.

\begin{table}
\begin{center}
\begin{small}
\begin{tabular}{|l|c|c|}
\hline
Database & Default & Maximum\\\hline
Actian Ingres 10.0/10S & S & S\\
Aerospike & RC & RC\\
Akiban Persistit & SI & SI\\
Clustrix CLX 4100 & RR & RR\\
Greenplum 4.1 & RC & S \\
IBM DB2 10 for z/OS & CS & S\\
IBM Informix 11.50 & Depends & S\\
MySQL 5.6 & RR & S \\
MemSQL 1b & RC & RC\\
MS SQL Server 2012 & RC & S \\
NuoDB & CR & CR\\
Oracle 11g & RC & SI\\
Oracle Berkeley DB & S & S\\
Oracle Berkeley DB JE & RR & S\\
Postgres 9.2.2 & RC & S\\
SAP HANA & RC & SI\\
ScaleDB 1.02 & RC & RC\\
VoltDB & S & S\\
\hline
\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline

@@

It is unknown \textit{which} of these ACID (particularly isolation)
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically designed for a
single-node context and are, to the best of our knowledge, unavailable
due to a reliance on concurrency control mechanisms like locking that
are not resilient to partial failure
(Section~\ref{sec:evaluation}). Morever, we are not aware of any prior
literature that provides guidance as to the relationship between weak
isolation and high availability: prior work has examined the
relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a single
server~\cite{adya, ansicritique, gray-isolation} but never weak
isolation and high availability taken together.  A primary goal in the
remainder of this paper is to understand which models are
HAT-compliant.

### 2013-03-31 writeup/relatedwork.tex
PDB pass plus Alan comments

There has been a recent resurgence of interest in distributed
multi-object semantics, both in academia~\cite{kraska-s3, granola,
  gstore, mdcc, redblue, cops, eiger, walter,calvin, swift} and
industry~\cite{megastore, orleans, spanner}. As discussed in
Section~\ref{sec:modernacid}, classic ACID databases provide strong
semantics but their lock-based and traditional multi-versioned
implementations are unavailable in the presence of
partitions~\cite{bernstein-concurrency, bernstein-book,
  gray-isolation}. Notably, Google's Spanner provides strong one-copy
serializable transactions. While Spanner is highly specialized for
Google's read-heavy workload, it relies on two-phase commit and
two-phase locking for read/write transactions~\cite{spanner}. As we
have discussed, the latency penalties associated with this design
choice are fundamental to serializability. For users willing to
tolerate these costs, Spanner, or similar strongly consistent,
unavailable systems---including Calvin~\cite{calvin},
G-Store~\cite{gstore}, Gemini~\cite{redblue}, Generalized Snapshot
Isolation~\cite{generalizedsnapshot}, Granola~\cite{granola},
HBase~\cite{hbase}, MDCC~\cite{mdcc}, Megastore~\cite{megastore},
Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis}, and
Walter~\cite{walter}---are a reasonable choice. We note that many
replication techniques target variants of Snapshot
Isolation~\cite{middleware-db}.

@@

>>>>

There has been a recent resurgence of interest in distributed
multi-object semantics, both in academia~\cite{kraska-s3, granola,
  gstore, mdcc, redblue, cops, eiger, walter,calvin, swift} and
industry~\cite{megastore, orleans, spanner}. As discussed in
Section~\ref{sec:modernacid}, classic ACID databases provide strong
semantics but their lock-based and traditional multi-versioned
implementations are unavailable in the presence of
partitions~\cite{bernstein-concurrency, bernstein-book,
  gray-isolation}. Notably, Google's Spanner provides strong one-copy
serializable transactions. While Spanner is highly specialized for
Google's read-heavy workload, it relies on two-phase commit and
two-phase locking for read/write transactions~\cite{spanner}. As we
have discussed, the latency penalties associated with this design
choice are fundamental to serializability. For users willing to
tolerate these costs, Spanner, or similar strongly consistent,
unavailable systems---including Calvin~\cite{calvin},
G-Store~\cite{gstore}, Gemini~\cite{redblue}, Generalized Snapshot
Isolation~\cite{generalizedsnapshot}, Granola~\cite{granola},
HBase, MDCC~\cite{mdcc}, Megastore~\cite{megastore},
Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis},
Walter~\cite{walter}, and a range of snapshot isolation
techniques~\cite{middleware-db, kemme-snapshot,
  daudjee-snapshot}---are a reasonable choice.

@@

From Serializable to Causal Transactions for Collaborative Applications: Causal consistency with lost update
